---
title: "Installation Instructions"
author: "Riley Xin"
date: "2023-10-06"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installation Instructions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## General
This guide navigates through some quirks and considerations during the installation of the `rosaceAA` and `cmdstanr` packages, particularly those related to compiler configurations on different operating systems. While installations on the latest MacOS and Linux with Ubuntu have been smooth, we've noted a few issues for installing on the Linux CentOS distribution related to C++ configurations. Alternatively, you may want to use a Docker container with the image we provided to bypass those issues.


## Installing `cmdstanr`
Before installing `rosaceAA`, you will need to install `cmdstanr` first using the code below or refer to the official [website](https://mc-stan.org/cmdstanr/).
```{r install cmdstanr, eval=FALSE}
install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))

# Use cmdstanr to install CmdStan, this requires a working C++ toolchain and compiler
library(cmdstanr)
install_cmdstan(cores = 4) # The number of CPU cores on your device

# Check that installation is successful by checking the CmdStan version
cmdstan_version()
```


If you are downloading on a Linux platform and encounter errors, it is likely due to requirements for a C++ compiler since `stan` is built in C++. According to [stan-dev](https://github.com/stan-dev/stan/wiki/Coding-Style-and-Idioms#supported-cpp-versions-and-compilers), `stan` is tested on Linux with g++ 5 (a compiler for C++). Ensure that your system is equipped with a gcc version > 5 which includes g++. You can check the current version in the terminal using:
```{bash, eval=FALSE}
gcc --version
```


If gcc is not installed or if your version is below the required, you'll need to install or upgrade it, following guidelines specific to your Linux distribution. You may check the distribution information in the terminal using:
```{bash, eval=FALSE}
lsb_release -a
```


## Installing `impute`
If you encounter `xx is not available for this version of R`, or `xx is not available for rosace-aa` during the automatic installation, you may try download it manually. For example, the package `impute` is not available on CRAN and can be downloaded using the following: 
```{r, eval = FALSE}
if (!require("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install("impute")
```

## Problem not solved?
You may submit an issue [on the GitHub repo](https://github.com/pimentellab/rosace-aa/issues) or email <roserao@ucla.edu>.
