---
title: "Introduction to rosaceAA"
author: "Jingyou Rao"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to rosaceAA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE, results="hide"}
knitr::opts_chunk$set(
  tidy = FALSE, cache = FALSE,
  dev = "png", dpi = 300, fig.align = "center", out.widt1h = "100%",
  fig.width = 6, fig.height = 4,
  message = FALSE, error = FALSE, warning = TRUE
)
```	

## Installation

To install `rosace` start R and first install devtools by typing:

```{r eval=FALSE}
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}
```

and install `rosace` from GitHub by typing (requires an Internet connection):

```{r eval=FALSE}
devtools::install_github("pimentellab/rosace-aa")
```

If you have cloned the git repository locally, navigate to the `rosace-aa` directory and type:

```{r eval=FALSE}
devtools::install(".")
```

Next load `rosaceAA` with:

```{r}
library("rosaceAA")
```

## Example

#### Read count data

To explain how to use `rosaceAA` we provide an example based on the OCT1 drug cytotoxicity screen (https://www.biorxiv.org/content/10.1101/2023.06.06.543963v1.full). 

The screen has three replicates. The count matrix is stored in `data/oct1.rda`.

```{r}
data("oct1")
key <- "1SM73"
type <- "growth"
```

#### Create Rosace object

First, load the count file into the `Assay` object. 
Each replicate in the experiment will form an `Assay` and will share the same "key" (1SM73).

```{r}
assay1 <- CreateAssayObject(
  counts = as.matrix(oct1_rep1[2:ncol(oct1_rep1)]),
  var.names = oct1_rep1$hgvs,
  key = key, rep = 1, type = type
)
assay2 <- CreateAssayObject(
  counts = as.matrix(oct1_rep2[2:ncol(oct1_rep2)]),
  var.names = oct1_rep2$hgvs,
  key = key, rep = 2, type = type
)
assay3 <- CreateAssayObject(
  counts = as.matrix(oct1_rep3[2:ncol(oct1_rep3)]),
  var.names = oct1_rep3$hgvs,
  key = key, rep = 3, type = type
)
```

Next create a `Rosace` object by concatenating three `Assay` objects.

```{r}
rosace <- CreateRosaceObject(object = assay1)
rosace <- AddAssayData(object = rosace, assay = assay2)
rosace <- AddAssayData(object = rosace, assay = assay3)
GetAssayName(rosace)
```

#### Preprocessing

`CreateAssayObject` calls function `FilterData` to filter the variants with more than `na.rmax` of NAs and less than `min.count` total counts by default. But we might want to filter more variants later.

```{r}
rosace <- FilterData(rosace, key = key, na.rmax = 0.5, min.count = 20)
```

Then we will impute the NA data either by K-Nearest Neighbor method or fill with 0. 

```{r}
rosace <- ImputeData(rosace, key = key, impute.method = "knn", na.rmax = 0.5)
# rosace <- ImputeData(rosace, key = key, impute.method = "zero")
```

With a complete count matrix in the `Assay` object, we will normalize the data by either a list of wild-type variants or by the total count at the time point. We recommend using the wild-type normalization because it will align the wild-type variant score to be 0 for hypothesis testing.

```{r}
rosace <- NormalizeData(rosace,
  key = key,
  normalization.method = "wt",
  wt.var.names = c("_wt"), wt.rm = TRUE
)
# rosace <- NormalizeData(rosace, key = key, normalization.method = "total")  # Total count normalization
```

After the `Assay` objects are all normalized, we can integrate the three replicates into an `AssaySet` object stored in the `combined.assay` slot of the `Rosace` object.

```{r}
rosace <- IntegrateData(object = rosace, key = key)
GetAssaySetName(rosace)
```

#### Process variants' meta data (provide your own function)

Provide your own function for parsing "hgvs" into position, mutation, wildtype, and type of mutation. An example (for parsing OCT1 variant names) is provided below:

```{r}
library("dplyr")
colnames(rosace@var.data)
rosace@var.data <- rosace@var.data %>%
  mutate(
    tmp = substr(variants, 4, nchar(variants) - 1),
    position = as.numeric(gsub("[[:alpha:]]", "", tmp)),
    wildtype = substr(tmp, 1, 1),
    tmp = substr(tmp, 2, nchar(tmp)),
    mutation = gsub("[[:digit:]]", "", tmp)
  ) %>%
  dplyr::select(-tmp)

func_map <- function(wt, mut) {
  if (nchar(wt) == 0) {
    return("NA")
  }

  if (wt == mut) {
    return("synonymous")
  } else if (mut == "del") {
    return("deletion")
  } else {
    return("missense")
  }
}

rosace@var.data <- rosace@var.data %>%
  rowwise() %>%
  mutate(type = func_map(wildtype, mutation)) %>%
  ungroup()
head(rosace@var.data)
```

#### Run Rosace

There are a few options to run the `RosaceAA` model. For details of the three models, refer to [bioRxiv link](https://www.biorxiv.org/content/10.1101/2025.01.09.632281v1).
Model 1 (position score only)
1. `pos.col`: provide position column
2. `ctrl.col`, `ctrl.name`, `stop.col`, `stop.name`: provide control and/or stop column (only useful if position column is provided).
Model 2 (position score + global substitution score)
3. `wt.col`, `mut.col`, `aa.code`: provide wildtype and mutation column (only useful when position column is provided) as well as the amino acid coding (single or triple).
Model 3 (position score + substitution score * position-specific activation)
4. `pos.act`: whether to add position-specific activation term

See below for example runs of the three models.

For the purpose of this vignette, we have reduced the number of variants to 100 with brute force before running Rosace, since running on the full dataset might take an hour.

```{r message=FALSE}
## Only process 100 variants for the vignette, comment out the slicing for all data

## Process the first assay
rosace@assays$`1SM73_1`@norm.counts <- rosace@assays$`1SM73_1`@norm.counts[1:100, ]
rosace@assays$`1SM73_1`@norm.var.names <- rosace@assays$`1SM73_1`@norm.var.names[1:100]

## Or process the entire assay set instead
rosace@assay.sets$`1SM73`@raw.counts <- rosace@assay.sets$`1SM73`@raw.counts[1:100, ]
rosace@assay.sets$`1SM73`@combined.counts <- rosace@assay.sets$`1SM73`@combined.counts[1:100, ]
rosace@assay.sets$`1SM73`@var.names <- rosace@assay.sets$`1SM73`@var.names[1:100]
```

```{r message=FALSE}
## Model 0 example
# running on an Assay (one replicate)
rosace <- RunRosace(
  object = rosace,
  name = "1SM73_1",
  type = "Assay",
  savedir = "../tests/results/assay_model0/",
  install = TRUE
)
# running on an AssaySet (all three replicates)
rosace <- RunRosace(
  object = rosace,
  name = "1SM73",
  type = "AssaySet",
  savedir = "../tests/results/assayset_model0/",
  install = FALSE
)
```

```{r message=FALSE}
## Model 1 example
# running on an Assay (one replicate)
rosace <- RunRosace(
  object = rosace,
  name = "1SM73_1",
  type = "Assay",
  savedir = "../tests/results/assay_model1/",
  pos.col = "position", # must
  ctrl.col = "type", # optional
  ctrl.name = "synonymous", # optional
  stop.col = "type", # optional
  stop.name = "deletion", # optional
  install = FALSE
)
# running on an AssaySet (all three replicates)
rosace <- RunRosace(
  object = rosace,
  name = "1SM73",
  type = "AssaySet",
  savedir = "../tests/results/assayset_model1/",
  pos.col = "position", # must
  ctrl.col = "type", # optional
  ctrl.name = "synonymous", # optional
  stop.col = "type", # optional
  stop.name = "deletion", # optional
  install = FALSE
)
```

```{r message=FALSE}
## MODEL 2 example
# running on an Assay (one replicate)
rosace <- RunRosace(
  object = rosace,
  name = "1SM73_1",
  type = "Assay",
  savedir = "../tests/results/assayset_model2/",
  pos.col = "position",
  ctrl.col = "type",
  ctrl.name = "synonymous",
  stop.col = "type",
  stop.name = "deletion",
  wt.col = "wildtype", # must
  mut.col = "mutation", # must
  aa.code = "single", # must
  pos.act = FALSE,
  install = FALSE
)
# running on an AssaySet (all three replicates)
rosace <- RunRosace(
  object = rosace,
  name = "1SM73",
  type = "AssaySet",
  savedir = "../tests/results/assayset_model2/",
  pos.col = "position",
  ctrl.col = "type",
  ctrl.name = "synonymous",
  stop.col = "type",
  stop.name = "deletion",
  wt.col = "wildtype", # must
  mut.col = "mutation", # must
  aa.code = "single", # must
  pos.act = FALSE,
  install = FALSE
)
```


```{r message=FALSE}
## MODEL 3 example
# running on an Assay (one replicate)
rosace <- RunRosace(
  object = rosace,
  name = "1SM73_1",
  type = "Assay",
  savedir = "../tests/results/assayset_model3/",
  pos.col = "position",
  ctrl.col = "type",
  ctrl.name = "synonymous",
  stop.col = "type",
  stop.name = "deletion",
  wt.col = "wildtype",
  mut.col = "mutation",
  aa.code = "single",
  pos.act = TRUE, # must
  install = FALSE
)
# running on an AssaySet (all three replicates)
rosace <- RunRosace(
  object = rosace,
  name = "1SM73",
  type = "AssaySet",
  savedir = "../tests/results/assayset_model2/",
  pos.col = "position",
  ctrl.col = "type",
  ctrl.name = "synonymous",
  stop.col = "type",
  stop.name = "deletion",
  wt.col = "wildtype",
  mut.col = "mutation",
  aa.code = "single",
  pos.act = TRUE, # must
  install = FALSE
)
```

The Rosace results can be retrieved using the `OutputScore` function. First, check the name of your Rosace run:
```{r}
names(rosace@scores)
```


Then extract the scores data, which includes the variant information, functional score (`mean`), standard deviation (`sd`), and the local false sign discovery rate (`lfsr`) associated with the score.
```{r message=FALSE}
# Take Assayset as an example
scores.data.model0 <- OutputScore(rosace, pos.info = FALSE, name = "1SM73_ROSACE0")
scores.data.model1 <- OutputScore(rosace, pos.info = TRUE, name = "1SM73_ROSACE1")
scores.data.model2 <- OutputScore(rosace, pos.info = TRUE, blosum.info = TRUE, name = "1SM73_ROSACE2")
scores.data.model3 <- OutputScore(rosace, pos.info = TRUE, blosum.info = TRUE, pos.act.info = TRUE, name = "1SM73_ROSACE3")
```

If the model outputs position-level information (model1, model2, and model3), the scores would return a list of two data frames, one for variant-level scores (df_variant) and another one for position-level scores (df_position).
```{r eval=FALSE}
head(scores.data.model0)
head(scores.data.model1$df_variant)
head(scores.data.model1$df_position)
head(scores.data.model2$df_variant)
head(scores.data.model2$df_position)
head(scores.data.model3$df_variant)
head(scores.data.model3$df_position)
```

The functional score represents the slope of the linear regression applied to normalized counts across time points, which serves as a measure of cell growth. A positive score indicates the mutation has a gain-of-function (GOF) effect, while a negative score indicates a loss-of-function (LOF) effect. Similar to `Rosace`, the `RosaceAA` model learns a distribution of scores and reports the mean for each variant. 

The `lfsr`, or local false sign rate, offers a Bayesian perspective for estimating uncertainty in the sign of the score. For a negative mean score, `lfsr` estimates the probability it could truly be positive, and vice versa. 


<div style="height: 200px;"></div>


